# Data 

## Sources

Source: http://fantasyoverlord.com/FPL/History
Season stats for 2016 to 2021 season

## Cleaning / transformation

The code we used for data cleaning and transformation can be seen in the “combine_data.Rmd” file. Our data for each season is parsed across one file for each week, as such the first step in our transformation process was to compile all of the weeks and seasons into one dataset. 

The `import_columns’ list shows the variables from each weekly data set that we compiled. In lines 32-47, we loop through each file, retain the desired columns, and append the observations to a master dataframe. From the filename of each weekly data set, we added a match week (‘Week’) column, as well as distinguished the season (‘Year’). Lastly, we converted fields from strings to numerics for later calculations.




## Missing value analysis

```{r}
# Libraries to use
library(tidyverse)
library(ggplot2)
library(ggthemes)
library(tibble)
library(dplyr) 
library(mi)
```


```{r}
#EF: Need to change link when new dataset uploaded to git *(RC)****************
df_orig <- read_csv("Data/main/master.csv")
# df_orig <- read_csv("https://raw.githubusercontent.com/ss16318/FantasyPremierLeaguePlayerValue/main/Data/main/master.csv")
```

```{r, fig.align = 'center'}
#reserved for showing 2019 missing weeks
df_missing_wk3637 = df_orig %>%
  select(Season, Week)

ggplot(df_missing_wk3637, aes(Season, factor(Week))) +
  stat_bin2d(aes(fill = after_stat(count)), binwidth = .999999) + 
  ggtitle("Visualization of Data Available by Season and Week") +
  theme(plot.title = element_text(hjust = 0.5)) + 
  theme_classic() +
  ylab("Week") +
  scale_fill_gradient(low = "white", high = "blue", limits=c(0,1000))
```

```{r}
# Missing Value (NA) Analysis
plot_missing2 <- function(x, percent = TRUE) {	
  na_count_all <- data.frame(is.na(x)) %>%	
    dplyr::group_by_all() %>%	
    dplyr::count(name = "count", sort = TRUE) %>%	
    dplyr::ungroup() %>%	
    tibble::rownames_to_column("pattern")	
  
  na_count_all <- na_count_all %>% 
    dplyr::mutate(pattern = factor(.data$pattern, levels = nrow(na_count_all):1))
  
  # count the number of columns with missing values; will be used later to determine if there's a "none missing" pattern	
  na_count_all <- na_count_all %>% 	
    dplyr::rowwise() %>%	
    dplyr::mutate(num_missing_cols = sum(dplyr::c_across(where(is.logical))))	
  
  # data frame for missing patterns bar chart	
  na_count_by_pattern <- na_count_all[,c("pattern", "count", "num_missing_cols")]
  na_count_by_pattern$none_missing <- ifelse(na_count_by_pattern$num_missing_cols == 0, TRUE, FALSE)

  # data frame for missing by column bar chart	
  na_count_by_column <- data.frame(is.na(x)) %>%	
    colSums() %>% 	
    sort(decreasing = TRUE) %>% 	
    tibble::enframe(name = "var", value = "count")	
  
  # tidy and sort na_count_all by column counts
  na_count_all_tidy <- na_count_all %>% 	
    tidyr::pivot_longer(where(is.logical), names_to = "variable") %>%	
    dplyr::mutate(variable = factor(.data$variable, levels = na_count_by_column$var))  %>% 	
    dplyr::mutate(none_missing = ifelse(.data$num_missing_cols == 0, TRUE, FALSE))	
  
  # main plot
  main_plot <- ggplot2::ggplot(na_count_all_tidy, ggplot2::aes(.data$variable, .data$pattern, fill = factor(.data$value), alpha = .data$none_missing)) +	
    ggplot2::geom_tile(color = "white") +	
    ggplot2::scale_fill_manual(values = c("grey70", "mediumpurple")) +	
    ggplot2::scale_alpha_manual(values = c(.7, 1)) +	
    ggplot2::ylab("missing pattern") +	
    ggplot2::guides(fill = "none", alpha = "none") +	
    ggplot2::theme_classic(12) + 
    ggplot2::theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) #MODIFICATION
    #ggplot2::theme(axis.text.y = element_text(angle = 90, vjust = 0.5, hjust=1)) #MODIFICATION
  
  # check for "none missing" pattern
  none_missing_pattern <- na_count_by_pattern %>%
    dplyr::filter(.data$none_missing) %>% dplyr::pull(.data$pattern)
  
  if (length(none_missing_pattern) > 0) {	
    main_plot <- main_plot +	
      ggplot2::annotate("text", x = (ncol(na_count_all)-2)/2,	
               y = nrow(na_count_all) + 1 - as.numeric(as.character(none_missing_pattern)),	
               label = "complete cases")	
  }	
  
  # margin plots
  denom <- ifelse(percent, nrow(x)/100, 1)
  
  missing_by_column_plot <- ggplot2::ggplot(na_count_by_column, ggplot2::aes(forcats::fct_inorder(.data$var), .data$count/denom)) +	
    ggplot2::geom_col(fill = "cornflowerblue", alpha = .7) +
    ggplot2::scale_y_continuous(expand = c(0, 0), n.breaks = 3) +	
    ggplot2::xlab("") +
    ggplot2::ylab(ifelse(percent, "% rows \n missing:", "num rows \n missing:")) +	
    ggplot2::theme_linedraw(12) + #modify	
    ggplot2::theme(panel.grid.major.x = ggplot2::element_blank(),	
          panel.grid.minor.x = ggplot2::element_blank()) +
    ggplot2::theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + #MODIFICATION
    ggplot2::theme(axis.text.y = element_text(size = 8)) #MODIFICATION
  
  missing_by_pattern_plot <- 
    ggplot2::ggplot(na_count_by_pattern, ggplot2::aes(.data$pattern, .data$count/denom, alpha = .data$none_missing)) +
    ggplot2::geom_col(fill = "cornflowerblue") +
    ggplot2::coord_flip() +
    ggplot2::scale_y_continuous(expand = c(0, 0), n.breaks = 3) +
    ggplot2::scale_alpha_manual(values = c(.7, 1)) +
    ggplot2::xlab("") +
    ggplot2::ylab(ifelse(percent, "% rows", "row count")) +
    ggplot2::guides(alpha = "none") +
    ggplot2::theme_linedraw(12) +
    ggplot2::theme(panel.grid.major.y = ggplot2::element_blank(), 
          panel.grid.minor.y = ggplot2::element_blank())
  
  if (percent) {	
    missing_by_column_plot <- missing_by_column_plot +
      ggplot2::scale_y_continuous(expand = c(0, 0), n.breaks = 5,
                         limits = c(0, 100))	
    missing_by_pattern_plot <- missing_by_pattern_plot +
      ggplot2::scale_y_continuous(expand = c(0, 0), n.breaks = 5,
                         limits = c(0, 100))	
  }	
  
  missing_by_column_plot + patchwork::plot_spacer() + 	
    main_plot + missing_by_pattern_plot + 	
    patchwork::plot_layout(widths = c(4, 1), heights = c(1, 1))
}
```

```{r}
#lv <- missing_data.frame(df_orig)
#levels(lv@patterns)
```

```{r, fig.align = 'center'}
plot_missing2(df_orig, percent = TRUE)
```



