# Data 

## Sources
As described in Section 2.2 Data Availability, all available CSV files of athletes' performance for any week in Seasons 2017, 2018, 2019, 2020, and 2021 can be found in the link below. All data downloaded for use can also be found in the Data folder.

Source: http://fantasyoverlord.com/FPL/History


## Cleaning / transformation

The code we used for data cleaning and transformation can be viewed in the “combine_data_v2.Rmd” file. Our data for each season is parsed across one file for each week, as such the first step in our transformation process was to compile all of the weeks and seasons into one dataset.

The `import_columns’ list shows the variables from each weekly data set that we compiled. In lines 32-47, we loop through each file, retain the desired columns, and append the observations to a master dataframe. From the filename of each weekly data set, we added a match week (‘Week’) column, as well as distinguished the season (‘Year’). Lastly, we converted field types of numerical fields from strings to numerics for use and later calculations.

The resulting dataframe is saved as a .CSV in Data/main with the name "master.csv."

[maybe talk about the filter cut and the point caculation here]

## Missing value analysis

```{r}
# Libraries to use
library(tidyverse)
library(ggplot2)
library(ggthemes)
library(tibble)
library(dplyr) 
library(mi)
```


```{r}
#EF: Need to change link when new dataset uploaded to git *(RC)****************
df_orig <- read_csv("Data/main/master.csv")
# df_orig <- read_csv("https://raw.githubusercontent.com/ss16318/FantasyPremierLeaguePlayerValue/main/Data/main/master.csv")
```

With all CSV files imported, merged, and verified for tidy, we know what an ideal dataset should look like – a dataframe with where observations of athlete performance should exists for every season between 2017 and 2021 for every week in those seasons. We then proceed to visualize the availability of our data over each season and week by showing tiles where they exist, so we can have an understanding of the distribution of observations over Season and Week. In the graph show, we can see that we can missing a few select seemingly random weeks in 2019 and 2020, but in 2019 we also observe that there is a missing pattern where every week after 32, with the exception of 35, is missing. Taking the data to context, one can realize that the missing data from the end of the 2019 season is due to the emergence of Covid-19. With knowledge of what weeks we are missing, we proceed forward in the analysis.

```{r, fig.align = 'center'}
# Missing Weeks Analysis
df_missing_wk3637 = df_orig %>%
  select(Season, Week)

ggplot(df_missing_wk3637, aes(Season, factor(Week))) +
  stat_bin2d(aes(fill = after_stat(count)), binwidth = .999999) + 
  ggtitle("Visualization of Data Available by Season and Week") +
  theme(plot.title = element_text(hjust = 0.5)) + 
  theme_classic() +
  ylab("Week") +
  scale_fill_gradient(low = "white", high = "blue", limits=c(0,1000))
```

Leveraging the source code of the “plot_missing” function for modification, the missing patterns in the dataset is plotted, showing the affected variables and the percent rows and percent rows missing. We find that about 90% of the data is complete and fully populated. Out of the last 10%, all of them are missing the entry for the first name of the players, and an indiscernible small number of rows are showing a missing pattern of missing YellowCards, CleanSheets, Assists, SelectedByPercent, OwnGoals, MinutesPlayed, and RedCards. Since the first names of the players present little value when the last names are present (since last names in the premier league are usually unique), the data we have on the weeks where available are effectively fully complete for the purpose of our analysis.

```{r}
# Missing Value (NA) Analysis
plot_missing2 <- function(x, percent = TRUE) {	
  na_count_all <- data.frame(is.na(x)) %>%	
    dplyr::group_by_all() %>%	
    dplyr::count(name = "count", sort = TRUE) %>%	
    dplyr::ungroup() %>%	
    tibble::rownames_to_column("pattern")	
  
  na_count_all <- na_count_all %>% 
    dplyr::mutate(pattern = factor(.data$pattern, levels = nrow(na_count_all):1))
  
  # count the number of columns with missing values; will be used later to determine if there's a "none missing" pattern	
  na_count_all <- na_count_all %>% 	
    dplyr::rowwise() %>%	
    dplyr::mutate(num_missing_cols = sum(dplyr::c_across(where(is.logical))))	
  
  # data frame for missing patterns bar chart	
  na_count_by_pattern <- na_count_all[,c("pattern", "count", "num_missing_cols")]
  na_count_by_pattern$none_missing <- ifelse(na_count_by_pattern$num_missing_cols == 0, TRUE, FALSE)

  # data frame for missing by column bar chart	
  na_count_by_column <- data.frame(is.na(x)) %>%	
    colSums() %>% 	
    sort(decreasing = TRUE) %>% 	
    tibble::enframe(name = "var", value = "count")	
  
  # tidy and sort na_count_all by column counts
  na_count_all_tidy <- na_count_all %>% 	
    tidyr::pivot_longer(where(is.logical), names_to = "variable") %>%	
    dplyr::mutate(variable = factor(.data$variable, levels = na_count_by_column$var))  %>% 	
    dplyr::mutate(none_missing = ifelse(.data$num_missing_cols == 0, TRUE, FALSE))	
  
  # main plot
  main_plot <- ggplot2::ggplot(na_count_all_tidy, ggplot2::aes(.data$variable, .data$pattern, fill = factor(.data$value), alpha = .data$none_missing)) +	
    ggplot2::geom_tile(color = "white") +	
    ggplot2::scale_fill_manual(values = c("grey70", "mediumpurple")) +	
    ggplot2::scale_alpha_manual(values = c(.7, 1)) +	
    ggplot2::ylab("missing pattern") +	
    ggplot2::guides(fill = "none", alpha = "none") +	
    ggplot2::theme_classic(12) + 
    ggplot2::theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) #MODIFICATION
    #ggplot2::theme(axis.text.y = element_text(angle = 90, vjust = 0.5, hjust=1)) #MODIFICATION
  
  # check for "none missing" pattern
  none_missing_pattern <- na_count_by_pattern %>%
    dplyr::filter(.data$none_missing) %>% dplyr::pull(.data$pattern)
  
  if (length(none_missing_pattern) > 0) {	
    main_plot <- main_plot +	
      ggplot2::annotate("text", x = (ncol(na_count_all)-2)/2,	
               y = nrow(na_count_all) + 1 - as.numeric(as.character(none_missing_pattern)),	
               label = "complete cases")	
  }	
  
  # margin plots
  denom <- ifelse(percent, nrow(x)/100, 1)
  
  missing_by_column_plot <- ggplot2::ggplot(na_count_by_column, ggplot2::aes(forcats::fct_inorder(.data$var), .data$count/denom)) +	
    ggplot2::geom_col(fill = "cornflowerblue", alpha = .7) +
    ggplot2::scale_y_continuous(expand = c(0, 0), n.breaks = 3) +	
    ggplot2::xlab("") +
    ggplot2::ylab(ifelse(percent, "% rows \n missing:", "num rows \n missing:")) +	
    ggplot2::theme_linedraw(12) + #modify	
    ggplot2::theme(panel.grid.major.x = ggplot2::element_blank(),	
          panel.grid.minor.x = ggplot2::element_blank()) +
    ggplot2::theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + #MODIFICATION
    ggplot2::theme(axis.text.y = element_text(size = 8)) #MODIFICATION
  
  missing_by_pattern_plot <- 
    ggplot2::ggplot(na_count_by_pattern, ggplot2::aes(.data$pattern, .data$count/denom, alpha = .data$none_missing)) +
    ggplot2::geom_col(fill = "cornflowerblue") +
    ggplot2::coord_flip() +
    ggplot2::scale_y_continuous(expand = c(0, 0), n.breaks = 3) +
    ggplot2::scale_alpha_manual(values = c(.7, 1)) +
    ggplot2::xlab("") +
    ggplot2::ylab(ifelse(percent, "% rows", "row count")) +
    ggplot2::guides(alpha = "none") +
    ggplot2::theme_linedraw(12) +
    ggplot2::theme(panel.grid.major.y = ggplot2::element_blank(), 
          panel.grid.minor.y = ggplot2::element_blank())
  
  if (percent) {	
    missing_by_column_plot <- missing_by_column_plot +
      ggplot2::scale_y_continuous(expand = c(0, 0), n.breaks = 5,
                         limits = c(0, 100))	
    missing_by_pattern_plot <- missing_by_pattern_plot +
      ggplot2::scale_y_continuous(expand = c(0, 0), n.breaks = 5,
                         limits = c(0, 100))	
  }	
  
  missing_by_column_plot + patchwork::plot_spacer() + 	
    main_plot + missing_by_pattern_plot + 	
    patchwork::plot_layout(widths = c(4, 1), heights = c(1, 1))
}
```

```{r}
#lv <- missing_data.frame(df_orig)
#levels(lv@patterns)
```

```{r, fig.align = 'center'}
plot_missing2(df_orig, percent = TRUE)
```



